A Gentle Introduction and Application of Feature-Based Clustering with Psychological Time Series

### 2023-06-12 manuscript.tex
Updates from Overleaf

Fortunately, past conceptual and empirical works in the experience sampling (ESM) literature have collected a number of meaningful aspects of psychological time series\footnote{In psychology, extensive longitudinal data collection methods are often referred to as experience sampling method (ESM), ecological momentary assessment (EMA), or ambulatory assessment (AA) studies. While the terms come for different conceptual backgrounds, they share a focus on collecting data over an extended period of time to capture people's behaviors and experiences as they vary over time and in response to different situations and events. In this article, we will use the experience sampling (ESM) term as it has the strongest footing within the clustering literature.}. Important aspects might include concerns over whether a symptom consistently stays at a certain level without much variability or whether some emotions develop together. For many of the most important developmental aspects, researchers have assembled numeric measures that capture these patterns. These summary statistics are often called ``dynamic measures'', ``principles of change'', or ``dynamic features'' of the psychological time series \citep{dejonckheere2019, kuppens2017, krone2018}. For most conceptualizations, the proposed number of aspects ranges from four to twelve key features \citep[][]{wang2006, dejonckheere2019}. Each of these statistics not only captures a distinct aspect of psychological time series, but also holds conceptual value --- inertia, for example, describes a resistance to change that can be indicative of psychological maladjustment \citep{kuppens2010} or a higher within-person variability can signal an erratic state \citep{myin-germeys2018}.

Yet, despite this rich diversity of meaningful time series features in psychology, most clustering of ESM data has only focused on a small and restrictive selection of time series characteristics. Thus far, the most common approach has been to cluster participants based on person-specific model parameters --- notably intercepts and slopes from vector autoregression models \citep[VAR; e.g.,][]{ariens2020, bulteel2016, stefanovic2022}. While such model parameters are familiar to researchers in the field, they are often restricted to autocorrelations and cross-lagged partial correlations \citep[e.g.,][]{bringmann2018c} --- only two of the many potentially important time series features. Additionally, the validity of model parameters is traditionally restricted by the assumptions of the particular statistical model. To take the common VAR model as an example, the model explicitly assumes that the time intervals between measurements are consistent (i.e., equidistant measurement assumption), it does not allow for missing values, and it assumes that means or variances do not change over time \citep[i.e., stationarity assumption;][]{lutkepohl2005}. These assumptions however stand incontrast to types of data researchers commonly collect to address important questions of erratic, and context-specific phenomena \citep[][]{myin-germeys2018, hamaker2017, kivela2022, helmich2020a}. Model parameters might, thus, not always accurately capture the time series and are often restricted in the features they capture.

@@

Recent efforts to address the shortcomings of model parameters for time series clustering have primarily proposed more complicated models, which either seek to relax specific assumptions \citep[e.g.,][]{denteuling2021} or include additional time series features \citep[e.g., see][]{krone2018}. In this manuscript, we instead, propose to directly cluster based on important and relevant features. Adequately named \textit{feature-based time series clustering}, such an approach has been a common procedure in digital phenotyping \citep[][]{loftus2022} and the broader machine learning literature \citep[][]{maharaj2019}. As such, the analysis has been applied to a variety of data, including analyses of astronomical, meteorological, and aviation pathways, biological and medical developments, as well as energy and finance patterns \citep{Aghabozorgi2015}. ESM APPLICATION. We argue that for psychological time series data, feature-based clustering offers a flexible variety, theory-based selection, fewer strict assumptions, easy and intuitive analysis, as well as meaningful psychological interpretability.

>>>>

Fortunately, past conceptual and empirical works in the experience sampling (ESM) literature have collected a number of meaningful aspects of psychological time series\footnote{In psychology, extensive longitudinal data collection methods are often referred to as experience sampling method (ESM), ecological momentary assessment (EMA), or ambulatory assessment (AA) studies. While the terms come from different conceptual backgrounds, they share a focus on collecting data over an extended period of time to capture people's behaviors and experiences as they vary over time and in response to different situations and events. In this article, we will use the experience sampling (ESM) term as it has the strongest footing within the clustering literature.}. Important aspects might include concerns over whether a symptom consistently stays at a certain level without much variability or whether some emotions develop together. For many of the most important developmental aspects, researchers have assembled numeric measures that capture these patterns. These summary statistics are often called ``dynamic measures'', ``principles of change'', or ``dynamic features'' of the psychological time series \citep{dejonckheere2019, kuppens2017, krone2018}. For most conceptualizations, the proposed number of aspects ranges from four to twelve key features \citep[][]{wang2006, dejonckheere2019}. Each of these statistics not only captures a distinct aspect of psychological time series but also holds conceptual value --- inertia, for example, describes a resistance to change that can be indicative of psychological maladjustment \citep{kuppens2010} or a higher within-person variability can signal an erratic state \citep{myin-germeys2018}.

Yet, despite this rich diversity of meaningful time series features in psychology, most clustering of ESM data has only focused on a small and restrictive selection of time series characteristics. Thus far, the most common approach has been to cluster participants based on person-specific model parameters --- notably intercepts and slopes from vector autoregression models \citep[VAR; e.g.,][]{ariens2020, bulteel2016, stefanovic2022}. While such model parameters are familiar to researchers in the field, they are often restricted to autocorrelations and cross-lagged partial correlations \citep[e.g.,][]{bringmann2018c} --- only two of the many potentially important time series features. Additionally, the validity of model parameters is traditionally restricted by the assumptions of the particular statistical model. To take the common VAR model as an example, the model explicitly assumes that the time intervals between measurements are consistent (i.e., equidistant measurement assumption), it does not allow for missing values, and it assumes that means or variances do not change over time \citep[i.e., stationarity assumption;][]{lutkepohl2005}. These assumptions however stand in contrast to types of data researchers commonly collect to address important questions of erratic, and context-specific phenomena \citep[][]{myin-germeys2018, hamaker2017, kivela2022, helmich2020a}. Model parameters might, thus, not always accurately capture the time series and are often restricted in the features they capture.

@@

Recent efforts to address the shortcomings of model parameters for time series clustering have primarily proposed more complicated models, which either seek to relax specific assumptions \citep[e.g.,][]{denteuling2021} or include additional time series features \citep[e.g., see][]{krone2018}. In this manuscript, we instead, propose to directly cluster based on important and relevant features. Adequately named \textit{feature-based time series clustering}, such an approach has been a common procedure in digital phenotyping \citep[][]{loftus2022} and the broader machine learning literature \citep[][]{maharaj2019}. As such, the analysis has been applied to a variety of data, including analyses of astronomical, meteorological, and aviation pathways, biological and medical developments, as well as energy and finance patterns \citep{Aghabozorgi2015}. RE-ADD ESM APPLICATIONS. We argue that for psychological time series data, feature-based clustering offers a flexible variety, theory-based selection, fewer strict assumptions, easy and intuitive analysis, as well as meaningful psychological interpretability.
